# Data transformation

## Introduction

```{r message=FALSE,cache=FALSE}
library("nycflights13")
library("tidyverse")
```

## Filter rows with `filter()`

### Exercise <span class="exercise-number">5.2.4.1</span> {.unnumbered .exercise}

<div class="question">
Find all flights that

1.  Had an arrival delay of two or more hours
1.  Flew to Houston (IAH or HOU)
1.  Were operated by United, American, or Delta
1.  Departed in summer (July, August, and September)
1.  Arrived more than two hours late, but didn’t leave late
1.  Were delayed by at least an hour, but made up over 30 minutes in flight
1.  Departed between midnight and 6am (inclusive)

</div>

<div class="answer">

The answer to each part follows.

1.  Using a couple of functions to knowledge, including `arrange()` and `%>%` (pipe). More on pipe
    later. Viewed the data `?` and `glimpse()` before constructing `filter()` functions.

    ```{r}
    data('flights')
    glimpse(flights)
    ?flights

    filter(flights, arr_delay >= 120) %>% arrange(arr_delay)
    ```

1.  Flew to Houston (IAH or HOU)
    ```{r}
    filter(flights, dest %in% c('IAH', 'HOU')) %>% arrange(dest)
    ```

1.  Were operated by United, American, or Delta
    ```{r}
    filter(flights, carrier %in% c('UA', 'AA', 'DL')) %>% arrange(carrier)
    ```

1.  Departed in summer (July, August, and September)
    ```{r}
    filter(flights, month >= 7, month <= 9)
    ```

1.  Arrived more than two hours late, but didn't leave late
    ```{r}
    filter(flights, dep_delay <= 0, arr_delay > 120)
    ```

1.  Were delayed by at least an hour, but made up over 30 minutes in flight
    ```{r}
    filter(flights, dep_delay >= 60 & dep_delay - arr_delay > 30) %>% 
      arrange(dep_delay, arr_delay)
    ```

1.  Departed between midnight and 6 a.m. (inclusive)
    ```{r}
    filter(flights, between(dep_time, 0, 600)) %>% arrange(dep_time)
    ```

</div>


